### 2019-04-25 papers/haskell19/4-generators.tex
Section 5

The remaining part of our random generator works analogously, except that this
time we only want to genrate terminal constructors, hence we use the |FreqOf'|
type family to compute the terminal generation frequency of each operand.
%
If any of |FreqOf' f| or |FreqOf' g| evaluates to zero, then it means that that
operand does not contain any terminal construction, and |frequency| will not
consider it when generating terminal values.

If it happens that both |FreqOf' f| and |FreqOf' g| compute to zero when
genrating a composite value, then it will produce a runtime error triggered by
the function |frequency|, as it does not have anything with a positive frequency
to generate.
%
This kind of exceptions will arise if we forget to tag at least a single
construction within a composite representation as terminal.
%
Fortunately, such runtime exceptions can be catched in compile time by defining
a type constraint that ensures we are trying to generate values using a
representation with a strictly positive terminal generation frequency---thus
containing at least a single terminal construction:

@@

Note how this primitive is also ambiguous in the type used for the
representation.
%
This allows us to use a visible type application to obtain values from the same
target type but generated using a different underlying representations.

So far we have seen how to represent and generate values for a target data type
by combining a wide selection of structural constructions, as well as a series
of type-level combinators to encode the generation behavior.
%
The next section refines our type-level machinery in order to export an elegant
domain specific language for specifying composable random generators.

>>>>

The remaining part of our random generator works analogously, except that this
time we only want to generate terminal constructions, hence we use the |FreqOf'|
type family to compute the terminal generation frequency of each operand.
%
If any of |FreqOf' f| or |FreqOf' g| evaluates to zero, then it means that that
operand does not contain any terminal constructions, and |frequency| will not
consider it when generating terminal values.

If it happens that both |FreqOf' f| and |FreqOf' g| compute to zero when
genrating a composite value, then it will produce a runtime error triggered by
the function |frequency|, as it does not have anything with a positive frequency
to generate.
%
This kind of exceptions will arise for examplke if we forget to tag at least a
single construction within a composite representation as terminal.
%
Fortunately, such runtime exceptions can be catched in compile-time by defining
a type constraint that ensures we are trying to generate values using a
representation with a strictly positive terminal generation frequency---thus
containing at least a single terminal construction:

@@

Note how this primitive is also ambiguous in the type used for the
representation.
%
This allows us to use a visible type application to obtain values from the same
target type but generated using a different underlying representations.
%
For instance, we can obtain two different concrete generators for our |Html|
type simply by changing its representation type as follows:

\begin{code}
genHtmlPats :: BGen Html
genHtmlPats = genEval at Html_simplify

genValidHtml :: BGen Html
genValidHtml = genEval  at ValidHtml
\end{code}

Where |Html_simplify| and |ValidHtml| are the types defined at the end of
Section \ref{sec:representation}.

% --------------------------------------
% Closing
%
So far we have seen how to represent and generate values for a target data type
by combining a wide selection of structural constructions, as well as a series
of type-level combinators to encode the generation behavior.
%
The next section refines our type-level machinery in order to export an elegant
domain specific language for specifying composable random generators.

### 2019-04-25 papers/haskell19/5-typelevel.tex
Section 5

@@

\begin{code}
instance BArbitrary Html where
  bounded = genEval at Html_Spec

instance Arbitrary Html where
  arbitrary = sized bounded
\end{code}

\begin{code}
type family Con (c :: Symbol)             :: * -> *
type family Pat (p :: Symbol) (n :: Nat)  :: * -> *
type family Fun (f :: Symbol)             :: * -> *
\end{code}

\begin{code}
type instance Con "Text"        = Con_Text
type instance Pat "simplify" 1  = Pat_simplify_1
type instance Fun "<+>"         = Fun_cat
\end{code}

@@

@@

\begin{code}
instance Algebra (Con_Leaf a) (Tree a) where ...
instance Algebra (Con_Node a) (Tree a) where ...
\end{code}

\begin{code}
instance Arbitrary a => BArbitrary1 (Con_Leaf a) where ...
instance Arbitrary a => BArbitrary1 (Con_Node a) where ...
\end{code}

@@

@@

\begin{code}
type Tree_Spec
  =      Con "Leaf" otimes 2
  oplus  Con "Node" otimes 3
\end{code}

\begin{code}
type family (f :: * -> *) apply (a :: *) :: * -> *  where
  (Some f)      apply  a  = f a
  (f oplus g)   apply  a  = (f apply a) oplus (g apply a)
  (f otimes n)  apply  a  = (f apply a) otimes n
  (Term f)      apply  a  = Term (t apply a)
  f             apply  a  = f
\end{code}

\begin{code}
instance Arbitrary a => BArbitrary (Tree a) where
  bounded = genEval at (Tree_Spec apply a)
\end{code}

>>>>

% --------------------------------------
% Intro
%
This section introduces a set refinements to our basic language for describing
random generators, making them more flexible and robust in order to fit
real-world usage scenarios.

% --------------------------------------
% Naming conventions
%
To begin with, let us consider the name conventions we used when defining our
representation data types along this paper.
%
Since our aim is to automate this process as much as possible using Haskell's
meta-programming capabilities, the actual names used when deriving construction
representations will have to be generated in such a way that they are compliant
with Haskell syntax and are unique within our namespace.
%
This means that, in practice, type names like |Fun_cat| or |Pat_simplify_1| are
technically not valid Haskell data type names, thus they will have to be
synthesized as something like |Fun_lt_plus_gt_543| and |Pat_simplify_1_325|,
where the last sequence of numbers is automatically inserted by Template Haskell
to ensure uniqueness.

This naming convention results hard to use, specially if we consider that we do
not now the actual name types until they are synthesized in compile-time, due to
their unique suffixes.
%
Fortunately, it is easy to solve this problem again using some type-level
machinery.
%
Instead of imposing a naming convention in our representation synthesizer, we
can define a set of open type families to hide each kind of construction using
meaningful names:

\begin{code}
type family Con (c :: Symbol)             :: * -> *
type family Fun (f :: Symbol)             :: * -> *
type family Pat (p :: Symbol) (n :: Nat)  :: * -> *
\end{code}

Where |Symbol| is the kind of type-level strings in Haskell.
%
Then, our derivation process will synthesize each representation using unique
names, along with a type instance of the corresponding type family for the end
user to manipulate it.
%
For instance, along with the representations of |Con_Text|, |Fun_cat| and
|Pat_simplify_1|, we will automatically derive the following type instances:

\begin{code}
type instance Con "Text"        = Con_Text_123
type instance Fun "<+>"         = Fun_lt_plus_gt_543
type instance Pat "simplify" 1  = Pat_simplify_1_325
\end{code}

And the end user can simply refer to each particular construction by using these
synonyms, e.g., with representation types like |Con "Text" oplus' Fun "<+>"|.

Moreover, we can define a \emph{terminal} version of each kind of construction
to avoid having to explicitly wrap constructions using |Term| combinators:

This way, any quoted construction will considered to be terminal.
%
With these type-level manipulations, we are now able to write generation
specifications in a clear and concise way.
%
For instance, we can specify a complex random generator for |Html| values as
follows:

@@

And simply obtain a concrete generator by refering to this previous type when
using |genEval|:

\begin{code}
genHtml :: BGen Html
genHtml = genEval at Html_Spec
\end{code}

% --------------------------------------
% Parametric target data types
%
\subsection{Parametric Target Data Types }

So far we have seen how specify random generators for our simple self-contained
|Html| data type.
%
In practice, however, we are often required to write random generators for
parametric target data types as well.
%
Consider for example the following |Tree| data type definition encoding binary
trees with generic information of type |a| in the leaves:

@@

In order to represent its data constructors, we can follow the same recipe
presented in Section \ref{sec:representation}, but also parametrizing our
representations over the type variable |a| as well:

@@

Then, the rest of the machinery can be derived in the same way as before,
carrying this type parameter and including the appropriate |Arbitrary|
constraints all along the way:

\begin{code}
instance Algebra (Con_Leaf a) (Tree a) where ...
instance Algebra (Con_Node a) (Tree a) where ...

instance Arbitrary a => BArbitrary1 (Con_Leaf a) where ...
instance Arbitrary a => BArbitrary1 (Con_Node a) where ...
\end{code}

The biggest different appears when deriving the required type instances for the
|Con|, |Fun| and |Pat| synonyms.
%
Instead of carrying this type parameter in our generation specifications, we can
avoid it by hiding them using an existential type:

@@

In the previous definition, |Some| is a wrapper for a 2-parametric type that
hides the first type variable by using an explicit existential quantificator.
%
Note thus that the type parameter |a| does not appear at the left hand side of
|Some| in its defition.
%
Then, when deriving any |Con|, |Fun| or |Pat| type synonym, we will wrap each
construction representations using a |Some|:

@@

This way we can write generation specification for our |Tree| data type without
having to refer to its type parameter anywhere.
%
For instance:

\begin{code}
type Tree_Spec
  =      Con'  "Leaf"  otimes 2
  oplus  Con   "Node"  otimes 3
\end{code}

Instead, we defer handling this type parameter until we actually use it to
define a concrete generator.
%
For instance, we can write a concrete generator of |Tree Int| as follows:

\begin{code}
genIntTree :: BGen (Tree Int)
genIntTree = genEval at (Tree_Spec apply' Int)
\end{code}

Where |apply| is a closed type family that simply traverses our generation
specification, applying the |Int| type to every occurrence of |Some|:

\begin{code}
type family (f :: * -> *) apply' (a :: *) :: * -> * space where
  (Some f)         apply'  a  = f a
  (f oplus''  g)   apply'  a  = (f apply' a) oplus'' (g apply' a)
  (f otimes'' n)   apply'  a  = (f apply' a) otimes'' n
  (Term f)         apply'  a  = Term (t apply' a)
  f                apply'  a  = f
\end{code}

As a result, in our definition of |genIntTree| above, the |apply| operator will
reduce the type |(Tree_Spec apply'' Int)| to the following concrete type:

\begin{code}
(Term (Con_Leaf Int) otimes' 2) oplus' ((Con_Node Int) otimes' 3)
\end{code}

Along with our automated constructions derivation mechanism, the machinery
introduced in this section allows us to specify random generators using a
concise and flexible type-level domain specific language.
%
The next section evaluates our approach in terms of performance using a set of
case studies extracted from real-world Haskell libraries.
